---
title: "naive bayes"
author: "Metales Abdelkader"
date: '2019-12-19'
summary: '**Naive bayes** model based on a strong assumption that the features are **conditionally  independent** given the class label...'
output:
  blogdown::html_page:
    toc: true
    number_sections: true
    fig_width: 6
    dev: "svg"
tags:
- naive bayes
categories: R
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE,message=FALSE,error=FALSE)
```


```{css, echo=FALSE}
strong {
  color: Navy;
}

h1,h2, h3, h4 {
  font-size:28px;
  color:DarkBlue;
}
```

# Introduction

**Naive bayes** model based on a strong assumption that the features are **conditionally  independent** given the class label. Since this assumption is rarely when it is true, this model termed as **naive**. However, even this assumption is not satisfied the model still works very well (Kevin.P murphy  2012). Using this assumption we can define the class conditionall density as the product of one dimensional densities.

$$p(X|y=c,\theta)=\prod_{j=1}^Dp(x_j|y=c,\theta_{jc})$$

The possible one dimensional density for each feature depends on the type of the feature: 

* For real_valued features we can make use of gaussion distribution:

$$p(X|y=c,\theta)=\prod_{j=1}^D\mathcal N(\mu_{jc}|y=c,\sigma_{jc}^2)$$

* For binary feature we can use bernouli distribution:

$$p(X|y=c,\theta)=\prod_{j=1}^DBer(x_j|\mu_{jc})$$

* For categorical feature we can make use of multinouli distribution:

$$p(X|y=c,\theta)=\prod_{j=1}^DCat(x_j|\mu_{jc})$$


For data that has features of different types we can use a mixture product of the above distributions, and this is what we will do in this paper.  

# Data preparation

The data that we will use here is [uploaded from kaggle website](https://www.kaggle.com/johnsmith88/heart-disease-dataset), which is about heart disease.
let us start by calling the packages needed and the data, then we give an appropriate name to the first column 
  

```{r}
library(tidyverse)
library(caret)
mydata<-read.csv("heart.csv",header = TRUE)
names(mydata)[1]<-"age"
glimpse(mydata)

```


the **target** variable indicates whether a patient has the disease or not based on the following features:

* age.
* sex: 1=male,0=female
* cp : chest pain type.
* trestbps :  resting blood pressure.
* chol: serum cholestoral.
* fbs : fasting blood sugar.
* restecg : resting electrocardiographic results.
* thalach : maximum heart rate achieved
* exang : exercise induced angina. 
* oldpeak : ST depression induced by exercise relative to rest.
* slope : the slope of the peak exercise ST segment.
* ca : number of major vessels colored by flourosopy.
* thal : it is not well defined from the data source.
* target: have heart disease or not.




The most intuitive thing by which we start  our analysis  is by getting the summary of this data to check the range, the five quantiles, and the existance or not of missing values for each feature.  


```{r}
summary(mydata)
```


 After inspecting the features we see that Some variables should be treated as factors rather than numerics such as **sex**, **cp**, **fbs**, **restecg**, **exange**, **slope**, **ca**, **thal**, and the **target** variable, hence they will be converted to factor type as follows: 
 
 
```{r}
mydata<-mydata %>%
  mutate_at(c(2,3,6,7,9,11,12,13,14),funs(as.factor))
summary(mydata)
```
 
 
 
 In practice It is very usefull to inspect (by traditional statistic test such as **kisq** or correlation coefficient) the relationships between the target variable and each of the potential explanatory variables before building any model, doing so we can tell apart the relevant variables from the irrelvant ones and hence which of which should include in our model. 
Another important issue with factors is that when spliting the data between training set and testing set some factor level can be missing in one set if the the number of casses for that level is too small.    
let's check if all the factor levels contribute on each target variable level.
 

```{r}
xtabs(~target+sex,data=mydata)
xtabs(~target+cp,data=mydata)
xtabs(~target+fbs,data=mydata)
xtabs(~target+restecg,data=mydata)
xtabs(~target+exang,data=mydata)
xtabs(~target+slope,data=mydata)
xtabs(~target+ca,data=mydata)
xtabs(~target+thal,data=mydata)
```
 
 
 
As we see the **restecg**,**ca** and **thal** variables have values less than the threshold of 5 casses required, so if we split the data between training set and test set the level **2** of the **restecg** variable will not be found in one of the sets since we have only one case. Therfore we should remove these variables from the model.

```{r}
mydata<-mydata[,-c(7,12,13)]
glimpse(mydata)
```



Before training our model, we can get a vague insight about the predictors that have some importance for the prediction of the dependent variable.  

Let's plot the relationships between the target variabl and the other features.


```{r}
ggplot(mydata,aes(sex,target,color=target))+
  geom_jitter()
```


If we look only at the red points (healthy patients) we can wrongly interpret that females are less healthy than males. This is because we do not take into account that we have imbalanced number of each sex level (96 females , 207 males). in contrast, if we look only at females we can say that a particular female are more likely to have the disease than not.      



```{r}
ggplot(mydata,aes(cp,fill= target))+
  geom_bar(stat = "count",position = "dodge")
```



From this plot we can conclude that if the patient does not have any chest pain he/she will be highly unlikely to get the disease, otherwise for any chest type the patient will be more likely to be pathologique by this disease. we can expect therfore that this predictor will have a significant importance on the training model.

```{r}
ggplot(mydata, aes(age,fill=target))+
  geom_density(alpha=.5)
```


Even there exist a large amount of overlapping between the two densities which may violate the independence assumption, it still exist some difference since these are drawn from the sample not the from the true distributions. However, we do not care much about it since we will evaluate the resulted model by using the testing set.  
we can also check this assumption with the corralation matrix. 


```{r}
library(psych)
pairs.panels(mydata[,-11])
```


AS we see all the correlations are less than 50% so we can go ahead and train our model.


# Data partition


we take out 80% of the data to use as training set and the rest will be put aside to evaluate the model performance.  


```{r}
set.seed(1234)
index<-createDataPartition(mydata$target, p=.8,list=FALSE)
train<-mydata[index,]
test<-mydata[-index,]
```


# Model training 

Note: for this model we do not need to set seed because this model uses known densities for the predictors and does not use any random method.

```{r}
library(naivebayes)
modelnv<-naive_bayes(target~.,data=train)
modelnv

```

As we see each predictor is treated depending on its type, gaussion distribution for numeric variables, bernouli distribution for binary variables and multinouli distribution for categorical variables.   

all the informations about this model can be extracted using the function **attributes**.  


```{r}
attributes(modelnv)
```

we can visualize the above reults with the fuction plot that provides us by plot the distribution of each features, densities for numeric features and bars for factors.  .

```{r}
plot(modelnv)
```




# Model evaluation

We can check the accuracy of the training data of this model using the confusion matrix.

```{r}
pred<-predict(modelnv,train)
confusionMatrix(pred,train$target)
```

The accuracy rate of the training set is about 79.84%.
as expected the specificity rate (81.82%) for class 1 is much larger than the snesitivity rate (77.48) for class 0. This is reflectd by the fact that we have larger number of class 1 than class 0.

```{r}
print(prop.table(table(train$target)),digits = 2)
      
```


The reliable evaluation is that based on the unseen testing data rather than the training data.  


```{r}
pred<-predict(modelnv,test)
confusionMatrix(pred,test$target)
```

The accuracy rate of the test set now is about 75%, may be due to overfitting problem, or this kind of model is not suitable for this data. 

# Model fine-tuning:

In order to increase the model performance we can try another set of hyperparameters. Naive bayes model has different kernels and by default the usekernel argument  is set to be **FALSE** which allows the use of the gaussion distriburtion for the numeric variables,if **TRUE** the kernel density estimation applies instead. Let's turn it to be **TRUE**  and see what will happen for the test accuracy rate.


```{r}
modelnv1<-naive_bayes(target~.,data=train,
                      usekernel = TRUE)
pred<-predict(modelnv1,test)
confusionMatrix(pred,test$target)

```

After using the kernel estimation we have obtained a slight improvement for the accuracy rate which is now about 76%.

Another way to improve the model is to try to preprocess the data, especailly for numeric when we standardize them they would follow the normal distribution.

```{r}
modelnv2<-train(target~., data=train,
                method="naive_bayes",
                preProc=c("center","scale"))
modelnv2
```

As we see we get better accuracy rate with the gaussion distribution 78.48% (when usekernel=FALSE) than with the kernel estimation 78.48%.


Let's use the test set:

```{r}
pred<-predict(modelnv2,test)
confusionMatrix(pred,test$target)


```

 We have another slight improvment with accuracy rate **78.33** after scaling the data. 
 
# Conclusion
 
Naive Bayes model is the most widely used model in the classical machine learning models, especially with features that are originally normally distributed or after transformation. However, compared to the bagged or boosted models like random forest exgboost models, or compared to deep learning models it is quite less attractive.   
 